### 2024-05-06 Paper/paper.tex
Updates from Overleaf

    \begin{center}
    \large
    \textbf{This is a preliminary draft. Please do not cite or circulate.}
	\end{center}

@@

    \begin{abstract}
		{\small
        In many industries, nonprofits make up the majority of firms in the market. Understanding the objectives and behaviors of these firms is important when creating incentives. However, nonprofit firms have goals and missions that make understanding their objectives difficult. For hospitals in particular, there is mixed evidence on whether nonprofit hospitals differ meaningfully from for-profit hospitals. In this paper, I investigate whether leadership team composition of nonprofit hospitals affects how similarly they behave to for-profits. I use a novel data set on nonprofit hospital executives in the US from 2009-2015 to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage two pay-for-performance initiatives in the US enacted in 2012, and estimate the difference in response between for-profits, nonprofits, and nonprofits with clinically trained executives. I find that while readmissions decreased on average for all types of hospitals, the response was larger for for-profits and nonprofits without clinically trained executives. This is consistent with theoretical predictions under the assumption that nonprofits with clinically trained executives put greater weight on societal outcomes than those without clinically trained executives.
		} 
	\end{abstract}

@@

  The objectives and behaviors of firms that are not classic for-profits are not easily determined, and an extensive literature has sought to understand these types of firms. Nonprofits are thought to gain utility not solely from monetary profit, but also through accomplishing societal benefit. Researchers have speculated a variety of objective functions that could motivate nonprofits such as maximizing prestige, income, or a quality/quantity trade-off (\cite{steinberg1986revealed}). Empirical contributions to this literature center around investigating behaviors that differ between for-profit, private nonprofit, and government firms (\cite{sloan2000not}). In this paper, I investigate whether differences in observed behavior between for-profit and nonprofit hospitals vary differentially by nonprofit leadership team composition.

  Hospitals have been a common focus when seeking to understand nonprofit firms. This is because, first, there are different hospital ownership types within the industry, giving researchers the ability to compare the behaviors of nonprofits with that of for-profits. Second, private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up around 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Thus, nonprofit behavior in the hospital industry is relevant, and directly affects consumers of health care. However, it remains unclear what objective function drives nonprofit hospital behavior, and whether nonprofit hospitals are meaningfully different from for-profits (\cite{sloan2000not}; \cite{erus2002inferring}; \cite{deneffe2002not}; \cite{horwitz2009hospital}).

  While the characterization of being for-profit or nonprofit is certainly important, there could be other characteristics that drive objectives and behavior that are not perfectly correlated with ownership status. Particularly, firm leadership varies widely across firms and potentially drives differences in objectives even within nonprofits. In large, publicly traded firms, executive characteristics are shown to be correlated with firm performance, indicating that there is more to firm behavior than for-profit or nonprofit status (\cite{bertrand2003managing}; \cite{matsa2013female}; \cite{ahern2012changing}). If hospital objectives differ in the weights placed on profit vs. societal benefit, then ownership status certainly plays a role in these weights, but characteristics of those managing the hospital could also play an important role. I contribute to our understanding of hospital and nonprofit behaviors by investigating whether occupational background, whether executives have clinical training, affects response to pay-for-performance incentives.

  Using publicly available tax forms, I construct a novel data set of nonprofit hospital executives from 2009-2015. I link hospitals in this data to characteristics in the American Hospital Association survey and Hospital Compare data. I leverage various pay-for-performance initiatives enacted in 2012 to investigate differential hospital responses between for-profits, nonprofits without clinically trained executives, and nonprofits with clinically trained executives. While executive teams change over time, I find no evidence that they change in response to the programs. This allows me to investigate the role of particular executives in two ways. First, I analyze the differential response to incentives of nonprofits who ever hire a clinically trained executive, as this might be a signal of underlying preferences. Second, I analyze the differential response of nonprofits who have a clinically trained executive in 2012 compared to those who have a clinically trained executive at other times, but not in 2012. These two analyses aim to disentangle the effect of the actual leadership team from the underlying preferences of the hospital.

  I estimate how the interaction between hospital type and post program enactment affects readmission and mortality rates among different groups. Under the assumption that hospitals would have continued to behave similarly in the absence of the programs, and that leadership teams do not change endogenously with the programs, I identify the differential response to pay-for-performance incentives by different hospital types. Among the hospitals in the data, hospitals with clinically trained executives are the minority, and differ on average from other hospitals along various dimensions. Thus, my preferred specification is synthetic difference-in-differences, to analyze a group of comparison hospitals with similar characteristics to treatment hospitals.\footnote{I also present results for the full sample, and with a pre-analysis matching procedure and find qualitatively similar results.}

   While all hospital types decrease readmission rates after the programs are enacted, for-profits and nonprofits without clinically trained executives decrease readmissions more than nonprofits without clinically trained executives. That is, nonprofits without clinically trained executives respond to incentives more similarly to for-profits than other nonprofits. To investigate the mechanisms by which hospitals decrease readmissions, I also conduct the analyses with uncompensated care and case mix index as outcomes. (results of this analysis here)

    This paper contributes to three strands of literature. First, it adds to our understanding of what underlies nonprofit hospital behavior. This literature is extensive, and a review of it is beyond the scope of this paper. To summarize, there are various aspects of nonprofit behavior and objectives for which the evidence of how much it differs from for-profits remains mixed: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). What underlies this is the need to understand the ways in which nonprofits make choices differently than for-profits. To this point, researchers have found that nonprofits do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011nonprofit} (\citeyear{chang2011nonprofit}), which looks at the effect of a cost shock on likelihood of shutting down and mix of profitable services. They find that nonprofits and for-profits are equally as likely to shut down after the shock, but only nonprofits adjust their mix of profitable services. Similarly, I investigate a change in incentives to hospitals, but I look at the differential impacts of leadership teams within nonprofits. This paper adds to our understanding of inputs into nonprofit hospital objective functions.

    Second, many researchers have documented a correlation between executive backgrounds, specifically for CEOs, and firm performance, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Female board members are associated with better oversight and more women executives (\cite{matsa2011chipping}; \cite{adams2009women}), but are negatively correlated with firm value (\cite{ahern2012changing}). Having more female executives is correlated with choices of female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions, while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}, \cite{adams2018director}; \cite{frydman2019rising}). Further, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}).

    Due to data limitations, many of these studies focus on large, publicly traded firms in the US. However, two studies focus on hospital performance in other countries. \citeauthor{janke2019impact} (\citeyear{janke2019impact}) uses data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). However, \citeauthor{otero2022managers} (\citeyear{otero2022managers}) investigates the role of CEOs in public hospitals and Chile, and finds an 8\% decrease in mortality rates for hospitals with top managers (\cite{otero2022managers}). I contribute to this literature by looking at the executive team as a whole rather than just CEOs, investigating a context which is not largely studied but is policy relevant, nonprofit US hospitals, and considering a unique executive characteristic pertinent to that context.

    Finally, I contribute to our understanding of how providers respond to pay-for-performance incentives in health care. The most in depth study of how hospitals respond to HRRP is \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}), who finds that hospitals decreased readmissions by 5\% and mortality rates by 2\% on average as a result of the program, confirming prior studies (\cite{mellor2017does}; \cite{ziedan2018essays}; \cite{ody2019decreases}; \cite{gupta2021impacts}). Importantly, around 40\% of the decrease in readmissions is due to selective patient practices. This raises important questions about how to incentivize improvement in quality of care without also incentivizing gamesmanship. Research on HVBP generally finds that the program had no effect on underlying hospital quality (\cite{us2015hospital}; \cite{norton2018moneyball}; \cite{friedson2019so}). This paper contributes to our understanding of how characteristics of hospitals affect response to pay-for-performance incentives.

@@

    Nonprofits are governed by a board of directors, whose role is to set high level goals and strategies, and provide general oversight. The board is not in charge of day-to-day operations, but the broad direction of the firm. They select executive team members to carry out day-to-day operations. While hospital executives typically do specialize by getting degree(s) in healthcare management or an MBA specific to healthcare, they do not often have clinical training. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive. An executive team usually consists of at least a CEO and CFO, but every nonprofit is different in exactly how they structure their executive team. Some nonprofits hire Chief Medical Officers, Chief Quality Officers, executive directors, vice presidents of various departments, etc. I consider all of these employees as part of the day-to-day management of the hospital and include them as a part of the executive team.

>>>>

@@

    \begin{abstract}
		{\small
        In many industries, not-for-profit (NFP) make up the majority of firms in the market. Understanding the objectives and behaviors of these firms is important when creating incentives. However, NFP firms have goals and missions that make understanding their objectives difficult. For hospitals in particular, there is mixed evidence on whether NFP hospitals differ meaningfully from for-profit (FP) hospitals. In this paper, I investigate whether leadership team composition of nonprofit hospitals affects how similarly they behave to for-profits. I use a novel data set on NFP hospital executives in the US (2009-2015) to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage two pay-for-performance initiatives in the US enacted in 2012, and estimate the difference in readmission and mortality rates afterwards between FP, NFP, and NFP with and without clinically trained executives. I find that NFPs without clinically trained executives behave more like FPs than other NFPs. This is consistent with theoretical predictions under the assumption that NFP hospitals with clinically trained executives value societal outcomes more than those without clinically trained executives.
		} 
	\end{abstract}

@@

  The objectives and behaviors of firms that are not classic for-profits are not easily determined, and an extensive literature has sought to understand these types of firms. Not-for-profits (NFP) are thought to gain utility not solely from monetary profit, but also through accomplishing societal benefit. Researchers have speculated a variety of objective functions that could motivate NFP firms such as maximizing prestige, income, or a quality/quantity trade-off (\cite{steinberg1986revealed}). Empirical contributions to this literature center around investigating behaviors that differ between for-profit (FP), private NFP, and government firms (\cite{sloan2000not}). In this paper, I investigate whether differences in observed behavior between for-profit and not-for-profit hospitals vary differentially by leadership team composition.

  Hospitals have been a common focus when seeking to understand NFP firms. This is because, first, there are different hospital ownership types within the industry, giving researchers the ability to compare the behaviors of NFPs and FPs. Second, private NFP hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, FPs make up around 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Thus, NFP behavior in the hospital industry is relevant, and directly affects consumers of health care. However, it remains unclear what objective function drives NFP hospital behavior, and whether NFP hospitals are meaningfully different from FP hospitals (\cite{sloan2000not}; \cite{erus2002inferring}; \cite{deneffe2002not}; \cite{horwitz2009hospital}).

  While the characterization of being for-profit or not-for-profit is certainly important, there could be other characteristics that affect hospital behavior that are not perfectly correlated with individual ownership status. Researchers have focused in particular on explaining the differences in NFP behavior by the market they exist again (citations). I argue that an observable and economically meaningful factor is the composition of leadership teams. Firm leadership varies widely across firms and potentially drives differences in objectives even within NFPs. In large, publicly traded firms, executive characteristics are shown to be correlated with firm performance, indicating that there is more to firm behavior than ownership status (\cite{bertrand2003managing}; \cite{matsa2013female}; \cite{ahern2012changing}). If hospital objectives differ in how they value monetary profit vs. gains to society, then ownership status certainly plays a role in these valuations, but the question remains as to whether management also plays an important role. I contribute to our understanding of hospital and NFP behaviors by investigating whether occupational background, specifically clinical training in executives, affects hospital response to pay-for-performance incentives.

  Using publicly available tax forms, I construct a novel data set of NFP hospital executives from 2009-2015. I link hospitals in this data to characteristics in the American Hospital Association (AHA) survey and Hospital Compare data. I leverage various pay-for-performance initiatives in 2012 to investigate differential hospital responses between for-profits, NFPs without clinically trained executives, and NFPs with clinically trained executives.

  I estimate how the interaction between hospital type and post program enactment affects readmission and mortality rates among different groups. Under the assumption that hospitals would have continued to behave similarly in the absence of the programs, and that leadership teams do not change endogenously with the programs, I identify the differential response to pay-for-performance incentives by different hospital types. Among the hospitals in the data, hospitals with clinically trained executives are the minority, and differ on average from other hospitals along various dimensions. Thus, my preferred specification is synthetic difference-in-differences, to analyze a group of comparison hospitals with similar characteristics to treatment hospitals.\footnote{I also present results for the full sample, and find qualitatively similar results.}

   While all hospital types decrease readmission rates after the programs are enacted, for-profits and not-for-profits without clinically trained executives decrease readmissions more than not-for-profits with clinically trained executives. That is, NFPs without clinically trained executives respond to incentives more similarly to FPs than other NFPs. I then consider whether this effect is due to clinically trained executives being a signal of underlying hospital preferences, or changing hospital preferences in a way that then affects behavior. I decompose the two by leveraging hospitals who change their propensity to hire a clinically trained executive at some point in the sample, which I find is not occurring endogenously with the change in incentives. The results of this analysis suggest that the difference in response is driven by changing hospital preferences, not as a signal of underlying preferences. While readmissions and mortality were targeted directly by the programs, I also consider uncompensated care and case mix index as outcomes, since re-allocation of resources and patient selectivity are potential mechanisms to changing hospital outcomes. (results of this analysis here)

    This paper contributes to three strands of literature. First, it adds to our understanding of what underlies NFP hospital behavior. There is mixed evidence on whether NFP behavior differs from FP behavior: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). Underlying hospital are hospital objectives. To this point, researchers have found that NFPs do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011nonprofit} (\citeyear{chang2011nonprofit}), which looks at the effect of a cost shock on likelihood of shutting down and mix of profitable services. They find that NFPs and FPs are equally as likely to shut down after the shock, but only NFPs adjust their mix of profitable services. Similarly, I investigate a change in incentives to hospitals, but I look at the differential impacts of leadership teams within nonprofits. This paper adds to our understanding of inputs into nonprofit hospital objective functions.

    Second, many researchers have documented a correlation between executive backgrounds, specifically for CEOs, and firm performance, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Female board members are associated with better oversight and more women executives (\cite{matsa2011chipping}; \cite{adams2009women}), but are negatively correlated with firm value (\cite{ahern2012changing}). Having more female executives is correlated with female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions, while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}; \cite{adams2018director}; \cite{frydman2019rising}). Further, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}).

    Due to data limitations, many of these studies focus on large, publicly traded firms in the US. \citeauthor{brickley2010board} (\citeyear{brickley2010board}) has the only study to my knowledge answering this question in the US NFP context. Using exogenous variation in expected Medicare profits, the authors find that having an internal board of directors increases CEO compensation, and having physician board members decreases public donations (\cite{brickley2010board}). Additionally, two studies focus on hospital performance in other countries. \citeauthor{janke2019impact} (\citeyear{janke2019impact}) uses data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). However, \citeauthor{otero2022managers} (\citeyear{otero2022managers}) investigates the role of CEOs in public hospitals and Chile, and finds an 8\% decrease in mortality rates for hospitals with top managers (\cite{otero2022managers}). I contribute to this literature by considering the executive team as a whole rather than just CEOs, by investigating a context which is not largely studied but is policy relevant, NFP US hospitals, and by leveraging a unique executive characteristic pertinent to that context.

    Finally, I contribute to our understanding of how providers respond to pay-for-performance incentives in health care. The most in depth study of how hospitals respond to HRRP is \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}), who finds that hospitals decreased readmissions by 5\% and mortality rates by 2\% on average as a result of the program, confirming prior studies (\cite{mellor2017does}; \cite{ziedan2018essays}; \cite{ody2019decreases}; \cite{gupta2021impacts}). Around 40\% of the decrease in readmissions is due to selective patient practices. Research on HVBP generally finds that the program had no effect on underlying hospital quality (\cite{us2015hospital}; \cite{norton2018moneyball}; \cite{friedson2019so}). This paper contributes to our understanding of how characteristics of hospitals affect response to pay-for-performance incentives.

@@

    NFPs are governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executive team members to carry out day-to-day operations. While hospital executives typically specialize by getting degree(s) in healthcare management or an MBA specific to healthcare, they do not often have clinical training. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive. An executive team usually consists of at least a CEO and CFO, but every NFP is different in exactly how they structure their executive team. Some NFPs hire Chief Medical Officers, Chief Quality Officers, executive directors, vice presidents of various departments, etc. I consider all of these employees as part of the day-to-day management of the hospital and include them as a part of the executive team.
